---
title: "(33) paper simulations"
author: "Christopher Perez"
date: "2026-01-21"
output: html_document
---

Simulation helper functions
```{r}
library(fdasrvf)
library(pracma)
library(cooltools)
library(foreach)
library(doParallel)
library(vegan)
library(patchwork)
```

This does not work for 2d object. This is only for 1d paramterized curves
```{r}
# Shape distance
# Depends on sampling points. Sample and true should have same number of points.
# We use the options to make it depend on scale (length should matter). 
library(fdasrvf)

L2dist <- function(sample_f, true_f){
  sqrt(sum((sample_f-true_f)^2))
}

# set scale = T for PCA line
geodesicDistance <-  function(sample,true,scale=F){
  # sample and true are shapes with same number of datapoints
  d <-calc_shape_dist(t(sample), t(true), mode="O", scale=scale, rotation=F)
  d$d
}

# For closed case. lambda infty is sample mean
L2distmean <- function(sample, sample_mean){
  sqrt(sum(sweep(sample, 2, sample_mean, "-")^2))
}
```


2D sphere


For each shape: ellipse, flower, spiky ball(?) and moon (show bad cases)
L2 distance  as N increases. Visualizations?

Lambda grid <- 10^(-13) to 10^(-3) , length .out =15?

Should I show L2 distance as iterations increase for a given sample size and lambda grid?



u we
Test to make sure it works

```{r}
set.seed(1)
N<-300
data <- flower2d3D(N, noise=0.02)
X <- data$X

plot_data_3D(X)

fit <- pmeClosed_2d(X, 10^seq(-8,-3, length.out=3), optimize_lambda = F, parallel=F)


Nsamp <- 1000
tseq <- fibonaccisphere(Nsamp)
plot_data_3D(tseq)
Xtrue <- flower2d3D(Nsamp, noise=0, t=tseq)$X_true
plot_data_3D(Xtrue)


Y <- rep(0, length(fit$spline_list))
for(i in 1:length(fit$spline_list))
{
  spline <- fit$spline_list[[i]]
  spline_sample <- spline(tseq)
  true_sample <- Xtrue
  
  Y[i] <- L2dist(spline_sample,  true_sample)*sqrt(4*pi/Nsamp)
  
}
      
      
plot(fit$lambda_list, Y, log='x')
lines(fit$lambda_list, Y)
  


plot_datafit_2d3D(X, fit$spline_list[[1]])
plot_datafit_2d3D(X, fit$spline_list[[2]])
plot_datafit_2d3D(X, fit$spline_list[[3]])
```

```{r}
plot_datafit_2d3D <- function(X, fit, mytitle="data and fit", col="seagreen"){
  surf <- genSurface(fit)
  
  plot_ly() %>%
    # Scatter points trace
    add_markers(x = X[, 1],y = X[, 2], z = X[, 3],
              type = "scatter3d",
              mode = "markers",
              marker = list(size = 2,opacity=0.7),
              name = "Data points"
    ) %>%
    # Overlay line trace
    add_surface(x=surf[[1]],y=surf[[2]], z=surf[[3]], opacity = 0.3,showscale=F,colorscale="viridis")|>
    layout(title = mytitle,
           scene = list(
             xaxis = list(title = "X"),
             yaxis = list(title = "Y"),
             zaxis = list(title = "Z")
           ))
}

plot_datafit_2d3D(X, fit$spline_list[[1]])

```

```{r}
Nsamp <- 1000
tseq <- fibonaccisphere(Nsamp)
Xtrue <- flower2d3D(Nsamp, noise=0, t=tseq)$X_true


Y <- rep(0, length(fit$spline_list))
for(i in 1:length(fit$spline_list))
{
  spline <- fit$spline_list[[i]]
  spline_sample <- spline(tseq)
  true_sample <- Xtrue
  
  Y[i] <- L2dist(spline_sample,  true_sample)*sqrt(4*pi/Nsamp)
  
}
Y
```

# 1d Circle


## Simulation code

```{r}
pmeClosed_1d_periodic_simulation <- function(shape_type,D, Ns,noise, lambdas, gridSize = 1000, max_iter = 20, err_tol=0.01, iso_k = 10,init_iso = F, L2dist_size = 1000, parallel=F,cl=NULL){
  X <- vector(mode="list", length=length(Ns))
  trues <- vector(mode="list", length=length(Ns))
  tcircles <- vector(mode="list", length=length(Ns))
  
  
  # the grid to evaluate at. this is also where we draw the data from.
  L2dist_grid <- seq(from=0,to=1-1e-6, length.out=L2dist_size)
  
  
  # Initialize data and true sample data
  if(shape_type == "ellipse" & D==2){
    
    for(j in seq_along(Ns)){
      
      X[[j]] <- circle1d2D(Ns[j], 1, 2, c(0,0), noise)$X
      trues[[j]] <- colMeans(X[[j]])#circle1d2D(L2dist_size, 1, 2, c(0,0), 0, t=L2dist_grid)$X
    }
    
  }else if(shape_type =="flower" & D==2){
    
    for(j in seq_along(Ns)){
      X[[j]] <- flower1d2D(Ns[j], 5, noise)$X
      trues[[j]] <- colMeans(X[[j]])##flower1d2D(L2dist_size,5, 0, t=L2dist_grid)$X
    }  
    
  }
  
  # Get initial projection indices for each dataset
  for(i in 1:length(Ns)){
    if(D == 2 & !init_iso){
      tcircles[[i]] <- project_circle(X[[i]])
    }else{
      iso2 <- isomap(dist(X[[i]]), ndim=2, k=iso_k)
      tcircles[[i]] <- project_circle(iso2$points)
    }
      
  }
  
  
  miniproj <- function(Xraw, newSpline){
    project_grid(Xraw, newSpline, gridSize)
  }
  
  # Setup parallel cluster
  stop_after <- FALSE
  if (is.null(cl) && parallel) {
    numCores <- parallel::detectCores() - 1
    cl <- makeCluster(numCores)
    registerDoParallel(cl)
    stop_after <- TRUE
  }
  
  # run parallel? T or F
  operator <- if(parallel) `%dopar%` else `%do%` #thanks chatgpt
  
  
  results <- vector(mode="list", length = length(Ns))
  
  # edit name to index by N
  names(results) <- paste0("N_",Ns)
  
  for(j in seq_along(Ns)){
    
    
    # Main algorithm loop
    results[[j]] <-operator( foreach(i = 1:length(lambdas), 
                                     .packages = c("stats", "pracma", "fdasrvf"),
                                     .export = c("spline1d_periodic",
                                                 "project_grid","project_circle", "project_optimize",
                                                 "spline_error","krt","krst", "L2distmean")), {
                                                   lambda <- lambdas[i]
                                                   splines <- list()
                                                   L2distances <- rep(NA, max_iter+1)
                                                   geodesicdistance <- rep(NA, max_iter+1)
                                                   L2index <- 1
                                                   
                                                   # Setup first spline fit
                                                   splines[[1]] <- spline1d_periodic(X[[j]],tcircles[[j]], lambdas[i])
                                                   
                                                   # Get first true manifold error
                                                   L2distances[1] <- L2distmean(splines[[1]](L2dist_grid), trues[[j]])
                                                   #geodesicdistance[1] <- geodesicDistance(splines[[1]](L2dist_grid), trues[[j]])
                                                   L2index <- L2index + 1
                                                   
                                                   # Setup first projection 
                                                   tt <- list() # list of t values
                                                   tt[[1]] <- miniproj(X[[j]],splines[[1]]) # evaluate first projection set
                                                   
                                                   # Repeat until error small
                                                   err <- err_tol + 1
                                                   iter <- 1
                                                   Diter <- rep(Inf, max_iter)
                                                   # Initial error
                                                   Diter[iter] <- spline_error(X[[j]], tt[[iter]], splines[[iter]])
                                                   
                                                   # Iteriate until error convergences
                                                   while(err > err_tol &  iter < max_iter){
                                                     iter <- iter+1
                                                     
                                                     #  Get newest spline fit
                                                     splines[[iter]] <- spline1d_periodic(X[[j]], tt[[iter-1]], lambdas[i])
                                                     # Get  true manifold error
                                                     L2distances[L2index] <- L2distmean(splines[[iter]](L2dist_grid), trues[[j]])
                                                     #geodesicdistance[L2index] <- geodesicDistance(splines[[iter]](L2dist_grid), trues[[j]])
                                                     L2index <- L2index + 1
                                                     
                                                     # Calculate new projection
                                                     tt[[iter]] <- miniproj(X[[j]],splines[[iter]]) # evaluate first projection set
                                                     
                                                     # Get newest error
                                                     Diter[iter] <- spline_error(X[[j]],tt[[iter]], splines[[iter]])
                                                     
                                                     # Compare old error
                                                     err <-abs((Diter[iter] - Diter[iter-1])/Diter[iter-1])
                                                   }
                                               
                                                   lambda_res <- list(
                                                     lambda = lambdas[i],
                                                    # spline_sequence = list(
                                                       splines = splines,
                                                       projections = tt,
                                                       max_iter = iter,
                                                       L2 = L2distances[1:iter]
                                                       #geodesic = geodesicdistance[1:iter]
                                                    # ),
                                                     #diagnostics = list(
                                                     #  L2 = L2distances[1:iter],
                                                    #   geodesic = geodesicdistance[1:iter]
                                                   #  )
                                                   )
                                                
                                                 })
    
    # Rename results j
    names(results[[j]]) <- paste0("lambda_", lambdas)
    results[[j]]$data <- X[[j]]
  }
  
    on.exit({
    if (stop_after) stopCluster(cl)
  }, add = TRUE)

  results
}

```



## Simulation Run
```{r}
Ns <- c(100,200,300)
noise <-0.01
lambdas <- c(1e-6,1e-3)
results <- pmeClosed_1d_periodic_simulation("ellipse",2,Ns, noise, lambdas, max_iter=5, L2dist_size=300)




library(tidyverse)
# from chatgpt
df_L2 <- map_dfr(seq_along(Ns), function(j) {
  map_dfr(seq_along(lambdas), function(i) {
    tibble(
      N      = Ns[j],
      lambda = lambdas[i],
      L2_last = tail(results[[j]][[i]]$L2, 1),
      geod_last = tail(results[[j]][[i]]$geodesic,1)
    )
  })
})

df_L2 %>% ggplot(aes(x =lambda, y=L2_last, col=factor(N))) + geom_point() + geom_line() + scale_x_log10()

ggplot(df_L2, aes(x = N, y = L2_last, col = factor(lambda))) +
  geom_line() +
  geom_point()


#df_L2 %>% ggplot(aes(x =lambda, y=geod, col=factor(N))) + geom_point() + geom_line() + scale_x_log10()

#ggplot(df_L2, aes(x = N, y = geod_last, col = factor(lambda))) +
#  geom_line() +
 # geom_point()



```


## Distance to Sample Mean
```{r}
library(latex2exp)
set.seed(1)
Ns <- c(100,200,500,1000)
noise <-0.03
lambdas <- 10^(-10:2)
results <- pmeClosed_1d_periodic_simulation("flower",2,Ns, noise, lambdas, max_iter=10, L2dist_size=300, parallel=T)
saveRDS(results, file = file.path("sim_results", "circle_mean.rds"))
results<-readRDS("sim_results/circle_mean.rds")


# viz
library(tidyverse)
df_L2 <- map_dfr(seq_along(Ns), function(j) {
  map_dfr(seq_along(lambdas), function(i) {
    tibble(
      N      = Ns[j],
      lambda = lambdas[i],
      L2_last = tail(results[[j]][[i]]$L2, 1)
      #geod_last = tail(results[[j]][[i]]$geodesic,1)
    )
  })
})

# geod by lambda
(Sample1d <- df_L2 %>% mutate(N=as.factor(N)) %>% ggplot(aes(x =lambda, y=L2_last, col=N,linetype=N)) + geom_point(aes(shape=N),size=2) + geom_line(linewidth=0.8) + scale_x_log10()  + theme_bw() + geom_hline(yintercept=0, linetype="dashed")+
  xlab("λ") + ylab("L2 Distance to Sample Mean") +
     theme(panel.grid=element_blank()))

# geod by N
#df_L2 %>% ggplot(aes(x =N, y=geod_last, col=factor(lambda))) + geom_point() + geom_line()
```



#1d Interval 

## asymptotics of line
```{r}
library(tidyverse)
set.seed(1)
n <- 100
t <- seq(0,1,length.out=n)
X <- t(cbind(t, sin(2*pi*t)))
X <- rotation2(pi/6)%*% X
X <- t(X) + rnorm(2*n, sd=0.03)
lambdas <- c(1e-15, 1e-6,1,10)

fit <- pmeClosed_1d_interval(X, lambdas, parallel=F, max_iter = 10, gridSize=2000)



S <- seq(from=0,to=1,length.out=1000)

plot(X, pch=16,cex=1)
cols <- rainbow(length(lambdas))
for(i in 1:length(lambdas))
{
  spl <- fit$spline_list[[i]]
  lines(spl(S),col=cols[i],lty=i,lwd=3)
}
legend("topright",legend = paste0("lambda ", lambdas),lty=1:length(fit),bty="n",col=cols,lwd=3)

# use chatgpt to generate ggplot version
## points
df_X <- data.frame(
  x = X[,1],
  y = X[,2]
)

## spline curves
df_splines <- do.call(
  rbind,
  lapply(seq_along(lambdas), function(i) {
    spl <- fit$spline_list[[i]]
    Y   <- spl(S)  # assumed m x 2
    
    data.frame(
      x = Y[,1],
      y = Y[,2],
      lambda = factor(lambdas[i])
    )
  })
)

ggplot() +
  geom_point(
    data = df_X,
    aes(x = x, y = y),
    size = 2
  ) +
  geom_path(
    data = df_splines,
    aes(x = x, y = y, color = lambda, linetype = lambda),
    linewidth =1
  ) + theme_bw() + xlab("X") + ylab("Y")
```
vs PCA

```{r}
pca <- prcomp(X, center=T)

pca1 <- pca$rotation[,1]
slope <- pca1[2]/pca1[1]
mu <- colMeans(X)
intr <- mu[2] - slope*mu[1]


ggplot() +
  geom_point(
    data = df_X,
    aes(x = x, y = y),
    size = 2
  ) + geom_path(data=df_splines %>% filter(lambda==10), aes(x=x,y=y, col="Smoothed Manifold"),size=1.5)+ geom_abline(aes(col="PCA") ,
slope=slope, intercept=intr,linetype="dashed", size=1.5)+scale_color_manual(
  values = c(
    "Smoothed Manifold" = "blue",
    "PCA" = "red"
  )
)
```
```{r}
rm(list = ls()[!sapply(ls(), function(x) is.function(get(x)))])
```

## Simulation code
```{r}
library(cooltools)
```

```{r}
pmeClosed_1d_interval_simulation <- function(shape_type, D, Ns, noise, lambdas, gridSize = 1000, max_iter = 10, err_tol=0.01, cl=NULL,parallel=F, k_init = 1, L2dist_size = 1000)
{
  # Begin actual code 
  X <- vector(mode="list", length=length(Ns))
  trues <- vector(mode="list", length=length(Ns))
  
  tinits <- vector(mode="list", length=length(Ns))
  
  # L2 dist initialize
  L2dist <- function(sample_f, true_f){
    sqrt(sum((sample_f-true_f)^2))
  }
  
  # the grid to evaluate at. this is also where we draw the data from.
  L2dist_grid <- seq(from=0,to=1-1e-6, length.out=L2dist_size)
  

  if(shape_type == "sine" & D==2){
    
    for(j in seq_along(Ns)){
      
      X[[j]] <- sine1d2D(Ns[j], rotation=pi/6,period=2*pi, noise=noise)$X
      trues[[j]] <- sine1d2D(L2dist_size, rotation=pi/6,period=2*pi, noise=noise)$X
    }
  }else if(shape_type =="PCA" & D==2){
    
    for(j in seq_along(Ns)){
      X[[j]] <- sine1d2D(Ns[j], rotation=pi/6,period=2*pi, noise=noise)$X
      
      # Get PCA line
      pca <- prcomp(X[[j]], center=T)
      pca1 <- pca$rotation[,1]
      slope <- pca1[2]/pca1[1]
      mu <- colMeans(X[[j]])
      intr <- mu[2] - slope*mu[1]
      
      trues[[j]] <- cbind(L2dist_grid, slope*L2dist_grid + intr)
    }  
    
  }
  use_scale <- shape_type=="PCA"
  
  # Initialize first projections 
  k_init<-1
  
  for(i in 1:length(Ns)){
    isores <- iso_try(X[[i]], k_init)
    iso <- isores$iso
    k <- isores$k
    print(paste0("k is ", toString(k)))
    tinits[[i]] <-normalize(iso$points)
  }


  # Setup parallel cluster
  # Setup cluster if not provided
  stop_after <- FALSE
  if (is.null(cl) && parallel) {
    numCores <- parallel::detectCores() - 1
    cl <- makeCluster(numCores)
    registerDoParallel(cl)
    stop_after <- TRUE
  }
  
  
  miniproj <- function(Xraw, newSpline){
    project_grid(Xraw, newSpline, gridSize)
  }
  # run parallel? T or F
  operator <- if(parallel) `%dopar%` else `%do%` #thanks chatgpt
  
  
  
  results <- vector(mode="list", length = length(Ns))
  names(results) <- paste0("N_",Ns)
  
  for(j in seq_along(Ns)){
    
    
    
    # Main algorithm loop
    results[[j]] <-operator( foreach(i = 1:length(lambdas), 
                                     .packages = c("stats", "pracma", "fdasrvf"),
                                     .export = c("spline1d_interval","project_grid", "project_optimize",
                                                 "spline_error","Rk","normalize", "L2dist", "geodesicDistance")), {
                                                   lambda <- lambdas[i]
                                                   splines <- list()
                                                   L2distances <- rep(NA, max_iter+1)
                                                   geodesicdistance <- rep(NA, max_iter+1)
                                                   L2index <- 1
                                                   
                                                   # Setup first spline fit
                                                   splines[[1]] <- spline1d_interval(X[[j]],tinits[[j]], lambdas[i])
                                                   
                                                   # Get first true manifold error
                                                   L2distances[1] <- L2dist(splines[[1]](L2dist_grid), trues[[j]])
                                                   geodesicdistance[1] <- geodesicDistance(splines[[1]](L2dist_grid), trues[[j]],use_scale)
                                                   L2index <- L2index + 1
                                                   
                                                   
                                                   
                                                   # Setup first projection 
                                                   tt <- list() # list of t values
                                                   tt[[1]] <-  normalize(miniproj(X[[j]],splines[[1]])) # evaluate first projection set
                                                   
                                                   # Repeat until error small
                                                   err <- err_tol + 1
                                                   iter <- 1
                                                   Diter <- rep(Inf, max_iter)
                                                   # Initial error
                                                   Diter[iter] <- spline_error(X[[j]], tt[[iter]], splines[[iter]])
                                                   
                                                   # Iteriate until error convergences
                                                   while(err > err_tol &  iter < max_iter){
                                                     iter <- iter+1
                                                     
                                                     #  Get newest spline fit
                                                     splines[[iter]] <- spline1d_interval(X[[j]], tt[[iter-1]], lambdas[i])
                                                   
                                                     # Get  true manifold error
                                                     L2distances[L2index] <- L2dist(splines[[iter]](L2dist_grid), trues[[j]])
                                                     geodesicdistance[L2index] <- geodesicDistance(splines[[iter]](L2dist_grid), trues[[j]],use_scale)
                                                     L2index <- L2index + 1
                                                     
                                                     # Calculate new projection
                                                     tt[[iter]] <-  normalize(miniproj(X[[j]],splines[[iter]]))
                                                     
                                                     # Get newest error
                                                     Diter[iter] <- spline_error(X[[j]],tt[[iter]], splines[[iter]])
                                                     
                                                     # Compare old error
                                                     err <-abs((Diter[iter] - Diter[iter-1])/Diter[iter-1])
                                                   }
                                                   
                                                   
                                                   lambda_res <- list(
                                                     lambda = lambdas[i],
                                                     # spline_sequence = list(
                                                     splines = splines,
                                                     projections = tt,
                                                     max_iter = iter,
                                                     L2 = L2distances[1:iter],
                                                     geodesic = geodesicdistance[1:iter]
                                                     # ),
                                                     #diagnostics = list(
                                                     #  L2 = L2distances[1:iter],
                                                     #   geodesic = geodesicdistance[1:iter]
                                                     #  )
                                                   )
                                                   
                                                 })
    
    # Rename results j
    names(results[[j]]) <- paste0("lambda_", lambdas)
    results[[j]]$data <- X[[j]]
  }
  
    on.exit({
    if (stop_after) stopCluster(cl)
  }, add = TRUE)

  results
}
```



## PCA Simulation
```{r}
library(purrr)
set.seed(1)
Ns <- c(100,200,500,1000)
noise <-0.03
lambdas <- 10^(-10:2)
results <- pmeClosed_1d_interval_simulation("PCA",2,Ns, noise, lambdas, max_iter=10, L2dist_size=2000, parallel=T)

#saveRDS(results, file = file.path("sim_results", "line_pca.rds"))
results<-readRDS("sim_results/line_pca.rds")


# viz
df_L2 <- map_dfr(seq_along(Ns), function(j) {
  map_dfr(seq_along(lambdas), function(i) {
    tibble(
      N      = Ns[j],
      lambda = lambdas[i],
      L2_last = tail(results[[j]][[i]]$L2, 1),
      geod_last = tail(results[[j]][[i]]$geodesic,1)
    )
  })
})

# geod by lambda
(PCA1d <- df_L2 %>% mutate(N=as.factor(N)) %>% ggplot(aes(x =lambda, y=geod_last, col=N,linetype=N)) + geom_point(aes(shape=N),size=2) + geom_line(linewidth=0.8) + scale_x_log10()  + theme_bw() +  geom_hline(yintercept=0, linetype="dashed")+
  xlab("λ") + ylab("Geodesic Distance to PCA Line") + theme(panel.grid=element_blank()))

# geod by N
#df_L2 %>% ggplot(aes(x =N, y=geod_last, col=factor(lambda))) + geom_point() + geom_line()

```



## Simulation run
```{r}
Ns <- c(100,200,500,1000)
noise <-0.05
lambdas <- c(1e-6,1e-3)
results <- pmeClosed_1d_interval_simulation("sine",2,Ns, noise, lambdas, max_iter=5, L2dist_size=2000)




library(tidyverse)
# from chatgpt
df_L2 <- map_dfr(seq_along(Ns), function(j) {
  map_dfr(seq_along(lambdas), function(i) {
    tibble(
      N      = Ns[j],
      lambda = lambdas[i],
      L2_last = tail(results[[j]][[i]]$L2, 1),
      geod_last = tail(results[[j]][[i]]$geodesic,1)
    )
  })
})

df_L2 %>% ggplot(aes(x =lambda, y=L2_last, col=factor(N))) + geom_point() + geom_line() + scale_x_log10()

ggplot(df_L2, aes(x = N, y = L2_last, col = factor(lambda))) +
  geom_line() +
  geom_point()


df_L2 %>% ggplot(aes(x =lambda, y=geod_last, col=factor(N))) + geom_point() + geom_line() + scale_x_log10()

ggplot(df_L2, aes(x = N, y = geod_last, col = factor(lambda))) +
  geom_line() +
  geom_point()



```



## Extractors
```{r}


# from chatgpt
df_L2 <- map_dfr(seq_along(Ns), function(j) {
  map_dfr(seq_along(lambdas), function(i) {
    tibble(
      N      = Ns[j],
      lambda = lambdas[i],
      L2_last = tail(results[[j]][[i]]$L2distances, 1),
      geod_last = tail(results[[j]][[i]]$geodesic_distances,1)
    )
  })
})

df_L2 %>% ggplot(aes(x =lambda, y=L2_last, col=factor(N))) + geom_point() + geom_line() + scale_x_log10()

ggplot(df_L2, aes(x = N, y = L2_last, col = factor(lambda))) +
  geom_line() +
  geom_point()


df_L2 %>% ggplot(aes(x =lambda, y=geod_last, col=factor(N))) + geom_point() + geom_line() + scale_x_log10() +
  geom_hline(aes(yintercept = 0),linetype="dashed")

ggplot(df_L2, aes(x = N, y = geod_last, col = factor(lambda))) +
  geom_line() +
  geom_point()

#S <- seq(from =0,to=1,length.out=1000)

#Nind <- 3
#lambdaind <- 5
#plot(X[[Nind]])

#lines(results[[Nind]][[lambdaind]]$splines[[results[[Nind]][[lambdaind]]$max_iter]](S))
```
```{r}
S <- seq(0,1,length.out=100)
plot(results[[3]][[4]]$splines[[results[[3]][[4]]$max_iter]](S),type="l",lty=2, lwd=2)
lines(trues[[3]], col="red")
legend("topright",legend=c("Manifold Fit", "PCA"), col=c("black","red"), lty=c(2,1))
```




# 2d Sphere


## Visualizations

```{r}
  library(plot3D)
  set.seed(1)
  lambdas <- 10^c(-12,-5.5, -3,-2.2)
  
  X1 <- flower2d3D(300, petals=6, noise=0.01)$X
  res <- pmeClosed_2d(X1,lambdas, parallel = F, optimize_lambda=F)
  
  
  X2 <- unifmoon2d3D(300, noise=0.01, bend_radius=2, bend=1.2*pi)$X
  res2 <- pmeClosed_2d(X2,lambdas, parallel = F, optimize_lambda=F, init_iso=T)

  

  
  # theta=20, phi =50,xlim=c(-0.7,0.7),ylim=NULL, zlim=c(-0.5,0.5)
# plotting CV by lambda

  png(filename="testplot.png", width=12, height=8, units="in",res=600)
  par(mar = c(1.5,2,1.5,2), mfrow=c(2,4), xpd=F)
  for(i in 1:length(lambdas)){
    plot_datafit_2d3D_static(X1, res$spline_list[[i]],theta=20, phi =50,bty="g",
                             title=paste0("λ = ", formatC(lambdas[i],format="e",digits=2)))
  }
  for(i in 1:length(lambdas)){
    plot_datafit_2d3D_static(X2, res2$spline_list[[i]],bty="g",theta=200, phi =40,
                            title=paste0("λ = ", formatC(lambdas[i],format="e",digits=2)))
  }
  
  
```



## Simulation code

```{r}
pmeClosed_2d_simulation <- function(shape_type, Ns, noise, lambdas, gridSize = 1000, max_iter = 10, err_tol=0.01, cl=NULL,parallel=F, iso_k=10, L2dist_size = 1000){
  
  # Begin actual code 
  X <- vector(mode="list", length=length(Ns))
  trues <- vector(mode="list", length=length(Ns))
  
  tspheres <- vector(mode="list", length=length(Ns))
  
  # L2 dist initialize
  L2dist <- function(sample_f, true_f){
    sqrt(sum((sample_f-true_f)^2))
  }
  
  # the grid to evaluate at. this is also where we draw the data from.
  L2dist_grid <- fibonaccisphere(n=L2dist_size)
  
  
  if(shape_type == "ellipse"){
    
    for(j in seq_along(Ns)){
      
      X[[j]] <- ellipse2d3D(Ns[j], c(1,2,1), noise=noise)
      trues[[j]] <- colMeans(X[[j]])
    }
  }else if(shape_type =="flower"){
    
    for(j in seq_along(Ns)){
      
      X[[j]] <- flower2d3D(Ns[j], petals=5, noise=noise)$X
      trues[[j]] <-colMeans(X[[j]])
      
    } 
  }
  
  
  # Initialize first projections 
  
  for(i in 1:length(Ns)){
    
    if(shape_type != "moon"){
      datac <- scale(X[[i]],scale=F)
      tspheres[[i]] <- datac/sqrt(rowSums(datac^2))
    }
    else{
      iso3 <- isomap(dist(X[[i]]), ndim=3, k=iso_k)
      t3 <- scale(iso3$points,scale=F)
      spheres[[i]] <-t3 / sqrt(rowSums(t3^2))
    }
  }


  # Setup parallel cluster
  # Setup cluster if not provided
  stop_after <- FALSE
  if (is.null(cl) && parallel) {
    numCores <- parallel::detectCores() - 1
    cl <- makeCluster(numCores)
    registerDoParallel(cl)
    stop_after <- TRUE
  }
  
  
  miniproj <- function(Xraw, newSpline){
    project_grid(Xraw, newSpline, gridSize)
  }
  # run parallel? T or F
  operator <- if(parallel) `%dopar%` else `%do%` #thanks chatgpt
  
  
  
  results <- vector(mode="list", length = length(Ns))
  names(results) <- paste0("N_",Ns)
  
  for(j in seq_along(Ns)){
    
    
    
    # Main algorithm loop
    results[[j]] <-operator( foreach(i = 1:length(lambdas), 
                                      .packages = c("stats", "pracma", "cooltools","Riemann"),
                     .export = c("spline2d","project_optimize2",
                                 "spline_error","qm","expr","Rpp","m","L2distmean")), {
                                                   lambda <- lambdas[i]
                                                   splines <- list()
                                                   L2distances <- rep(NA, max_iter+1)
                                                   geodesicdistance <- rep(NA, max_iter+1)
                                                   L2index <- 1
                                                   
                                                   # Setup first spline fit
                                                   splines[[1]] <- spline2d(X[[j]],tspheres[[j]], lambdas[i])
                                                   
                                                   # Get first true manifold error
                                                   L2distances[1] <- L2distmean(splines[[1]](L2dist_grid), trues[[j]])
                                                   #geodesicdistance[1] <- geodesicDistance(splines[[1]](L2dist_grid), trues[[j]],use_scale)
                                                   L2index <- L2index + 1
                                                   
                                                  # Setup first projection 
                                                   tt <- list() # list of t values
                                                   tt[[1]] <-  project_optimize2(X[[j]], splines[[1]], tspheres[[j]]) # evaluate first projection set
                                                   
                                                   # Repeat until error small
                                                   err <- err_tol + 1
                                                   iter <- 1
                                                   Diter <- rep(Inf, max_iter)
                                                   # Initial error
                                                   Diter[iter] <- spline_error(X[[j]],tt[[iter]], splines[[iter]])
                                                   
                                                   # Iteriate until error convergences
                                                   while(err > err_tol &  iter < max_iter){
                                                     iter <- iter+1
                                                     
                                                     #  Get newest spline fit
                                                     splines[[iter]] <- spline2d(X[[j]], tt[[iter-1]], lambdas[i])
                                                   
                                                     # Get  true manifold error
                                                     L2distances[L2index] <- L2distmean(splines[[iter]](L2dist_grid), trues[[j]])
                                                     #geodesicdistance[L2index] <- geodesicDistance(splines[[iter]](L2dist_grid), trues[[j]],use_scale)
                                                     L2index <- L2index + 1
                                                     
                                                     # Calculate new projectio
                                                     tt[[iter]] <-  project_optimize2(X[[j]], splines[[iter]], tt[[iter-1]])
                                      
                                                     # Get newest error
                                                     Diter[iter] <- spline_error(X[[j]],tt[[iter]], splines[[iter]])
                                                     
                                                     # Compare old error
                                                     err <-abs((Diter[iter] - Diter[iter-1])/Diter[iter-1])
                                                   }
                                                   
                                                   
                                                   lambda_res <- list(
                                                     lambda = lambdas[i],
                                                     # spline_sequence = list(
                                                     splines = splines,
                                                     projections = tt,
                                                     max_iter = iter,
                                                     L2 = L2distances[1:iter]
                                                    # geodesic = geodesicdistance[1:iter]
                                                     # ),
                                                     #diagnostics = list(
                                                     #  L2 = L2distances[1:iter],
                                                     #   geodesic = geodesicdistance[1:iter]
                                                     #  )
                                                   )
                                                   
                                                 })
    
    # Rename results j
    
    names(results[[j]]) <- paste0("lambda_", lambdas)
    results[[j]]$data <- X[[j]]
  }
  results
}
```

## Simulation run
```{r}
Ns <- c(100,200)
noise <-0.01
lambdas <- c(1e-6,1e-3)
results <- pmeClosed_2d_simulation("flower",Ns, noise, lambdas, max_iter=5, L2dist_size=300,parallel=F)

S <- fibonaccisphere(n=1000)
results$N_100$`lambda_1e-06`$max_iter
plot_surf(results$N_100$`lambda_1e-06`$splines[[1]])


plot_surf(results$N_200$lambda_0.001$splines[[2]])

plot_datafit_2d3D(results$N_100$data,results$N_100$`lambda_1e-06`$splines[[1]] )


```

## Distance to sample mean, sphere
```{r}
set.seed(1)
Ns <- c(100,200,500,1000)
noise <-0.03
lambdas <- 10^(-10:0)
results <- pmeClosed_2d_simulation("flower",Ns, noise, lambdas, max_iter=10, L2dist_size=1000, parallel=T)
saveRDS(results, file = file.path("sim_results", "sphere_mean.rds"))
results<-readRDS("sim_results/sphere_mean.rds")


# viz
library(tidyverse)
df_L2 <- map_dfr(seq_along(Ns), function(j) {
  map_dfr(seq_along(lambdas), function(i) {
    tibble(
      N      = Ns[j],
      lambda = lambdas[i],
      L2_last = tail(results[[j]][[i]]$L2, 1)
      #geod_last = tail(results[[j]][[i]]$geodesic,1)
    )
  })
})

# geod by lambda
(Sample2d <- df_L2 %>% mutate(N=as.factor(N)) %>% ggplot(aes(x =lambda, y=L2_last, col=N,linetype=N)) + geom_point(aes(shape=N),size=2) + geom_line(linewidth=0.8) + scale_x_log10()  + theme_bw() + geom_hline(yintercept=0, linetype="dashed")+
  xlab("λ") + ylab("L2 Distance to Sample Mean")  +theme(panel.grid=element_blank()) + ggtitle("(iii) 2-Sphere"))

# geod by N
#df_L2 %>% ggplot(aes(x =N, y=geod_last, col=factor(lambda))) + geom_point() + geom_line()
```

# PCA sim full
```{r}
#ibrary(latex2exp)
set.seed(1)
Ns <- c(100,200,500,1000)
noise <-0.03
lambdas1 <- 10^(-10:2)

lambdas3 <- 10^(-10:0)
#results1 <- pmeClosed_1d_periodic_simulation("flower",2,Ns, noise, lambdas1, max_iter=10, L2dist_size=300, parallel=T)
#saveRDS(results1, file = file.path("sim_results", "circle_mean.rds"))
#results2 <- pmeClosed_1d_interval_simulation("PCA",2,Ns, noise, lambdas1, max_iter=10, L2dist_size=2000, parallel=T)
#saveRDS(results2, file = file.path("sim_results", "line_pca.rds"))
#results3 <- pmeClosed_2d_simulation("flower",Ns, noise, lambdas3, max_iter=10, L2dist_size=1000, parallel=T)
#saveRDS(results3, file = file.path("sim_results", "sphere_mean.rds"))

results1<-readRDS("sim_results/circle_mean.rds")
results2<-readRDS("sim_results/line_pca.rds")
results3<-readRDS("sim_results/sphere_mean.rds")



# viz
library(tidyverse)
df_circ <- map_dfr(seq_along(Ns), function(j) {
  map_dfr(seq_along(lambdas1), function(i) {
    tibble(
      N      = Ns[j],
      lambda = lambdas1[i],
      L2_last = tail(results1[[j]][[i]]$L2, 1)
      #geod_last = tail(results[[j]][[i]]$geodesic,1)
    )
  })
})

# viz
df_line <- map_dfr(seq_along(Ns), function(j) {
  map_dfr(seq_along(lambdas1), function(i) {
    tibble(
      N      = Ns[j],
      lambda = lambdas1[i],
      L2_last = tail(results2[[j]][[i]]$L2, 1),
      geod_last = tail(results2[[j]][[i]]$geodesic,1)
    )
  })
})

# viz
library(tidyverse)
df_sph <- map_dfr(seq_along(Ns), function(j) {
  map_dfr(seq_along(lambdas3), function(i) {
    tibble(
      N      = Ns[j],
      lambda = lambdas3[i],
      L2_last = tail(results3[[j]][[i]]$L2, 1)
      #geod_last = tail(results[[j]][[i]]$geodesic,1)
    )
  })
})

leg <- list(theme(legend.background=element_rect(fill="white",color="black"), panel.grid=element_blank(),aspect.ratio=1))


# geod by lambda
(Sample1d <- df_circ %>% mutate(N=as.factor(N)) %>% ggplot(aes(x =lambda, y=L2_last, col=N,linetype=N)) + geom_point(aes(shape=N),size=2) + geom_line(linewidth=0.8) + scale_x_log10()  + theme_bw() + geom_hline(yintercept=0, linetype="dashed")+
  xlab("λ") + ylab("L2 Distance to Sample Mean") +
     theme(panel.grid=element_blank())  +  ggtitle("(ii) Circle")+leg)


# geod by lambda
(PCA1d <- df_line %>% mutate(N=as.factor(N)) %>% ggplot(aes(x =lambda, y=geod_last, col=N,linetype=N)) + geom_point(aes(shape=N),size=2) + geom_line(linewidth=0.8) + scale_x_log10()  + theme_bw() +  geom_hline(yintercept=0, linetype="dashed")+
  xlab("λ") + ylab("Geodesic Distance to PCA Line") + theme(panel.grid=element_blank()) +ggtitle("(i) Interval")+leg)

# geod by lambda
(Sample2d <- df_sph %>% mutate(N=as.factor(N)) %>% ggplot(aes(x =lambda, y=L2_last, col=N,linetype=N)) + geom_point(aes(shape=N),size=2) + geom_line(linewidth=0.8) + scale_x_log10()  + theme_bw() + geom_hline(yintercept=0, linetype="dashed")+
  xlab("λ") + ylab("L2 Distance to Sample Mean")  +theme(panel.grid=element_blank()) + ggtitle("(iii) 2-Sphere")+leg)


(myplt<-PCA1d + Sample1d + Sample2d + plot_layout(guides="collect") & theme(legend.position="bottom"))
ggsave("sim_results/dist_lambda_infty.png",width=12, height=4, dpi=600)


```



# Isomap unwound example

```{r}
library(scico)
set.seed(1)
# data set
n <- 300
X <- moon1d2D(n, 1, 2, 0.06)
colz <- rainbow(nrow(X))
colz <- scico(nrow(X),palette= "romaO")

# plot
par(mfrow=(c(1,2)))
plot(X, col =colz)
iso <- isomap(dist(X), ndim=2, k=10)
plot(iso$points,col =colz)


# pme
set.seed(1)

spl_naive_first <- pmeClosed_1d_periodic(X, c(1e-8), init_iso=F, max_iter=1)
spl_iso_first<- pmeClosed_1d_periodic(X, c(1e-8), init_iso=T,max_iter=1)

spl_naive <- pmeClosed_1d_periodic(X, c(1e-8), init_iso=F, max_iter=10)
spl_iso<- pmeClosed_1d_periodic(X, c(1e-8), init_iso=T,max_iter=10)

S <- seq(0,1,length.out=1000)
```

```{r}


par(mfrow=c(1,2))
xlim <- c(1.,0.9)*c(min(X[,1]), max(X[,1]))
ylim <-c(1.,0.9)*c(min(X[,2]), max(X[,2]))
plot(spl_naive_first$spline_list[[1]](S), asp=1, type="l", col="red", xlim=xlim,ylim=ylim)
lines(spl_iso_first$spline_list[[1]](S),asp=1, col="blue", xlim=xlim,ylim=ylim, lty=2)
plot(spl_naive$spline_list[[1]](S), asp=1, type="l", col="red", xlim=xlim,ylim=ylim, lwd=2)
lines(spl_iso$spline_list[[1]](S), asp=1, col="blue", xlim=xlim,ylim=ylim,lwd=2,lty=2)



#plot(spl_iso_first$spline_list[[1]](S), asp=1, type="l", col="red", xlim=xlim,ylim=ylim,lty=2)
#lines(spl_naive_first$spline_list[[1]](S),asp=1, col="blue", xlim=xlim,ylim=ylim)
#plot(spl_iso$spline_list[[1]](S), asp=1, type="l", col="red", xlim=xlim,ylim=ylim,lty=2, lwd=2)
#lines(spl_naive$spline_list[[1]](S), asp=1, col="blue", xlim=xlim,ylim=ylim,lwd=2)

#par(mfrow=c(1,2))
m <- matrix(1:3, nrow=1, byrow=T)
par(mfrow=c(1,3))
xlim <- c(1.,0.9)*c(min(X[,1]), max(X[,1]))
ylim <-c(1.,0.9)*c(min(X[,2]), max(X[,2]))
plot(spl_naive_first$spline_list[[1]](S), asp=1, type="l", col="red", xlim=xlim,ylim=ylim,lwd=2)
lines(spl_iso_first$spline_list[[1]](S),asp=1, col="blue", xlim=xlim,ylim=ylim, lty=2,lwd=2)
plot(spl_naive$spline_list[[1]](S), asp=1, type="l", col="red", xlim=xlim,ylim=ylim, lwd=2)
lines(spl_iso$spline_list[[1]](S), asp=1, col="blue", xlim=xlim,ylim=ylim,lwd=2,lty=2)
plot(1,type="n", axes=F,xlab="",ylab="")
legend(x="top", legend=c("ISOMAP into Circle", "Naive Circle"), col = c("red","blue"),lty=c(1,2),lwd=2)


```

```{r}
#gpt
nf  <- spl_naive_first$spline_list[[1]](S)
if1 <- spl_iso_first$spline_list[[1]](S)
nc  <- spl_naive$spline_list[[1]](S)
ic  <- spl_iso$spline_list[[1]](S)

df <- rbind(
  data.frame(s = S, x = nf[,1], y = nf[,2],
             Method = "Naive", Stage = "Initial"),

  data.frame(s = S, x = if1[,1], y = if1[,2],
             Method = "ISOMAP", Stage = "Initial"),

  data.frame(s = S, x = nc[,1], y = nc[,2],
             Method = "Naive", Stage = "Converged"),

  data.frame(s = S, x = ic[,1], y = ic[,2],
             Method = "ISOMAP", Stage = "Converged")
)
df$Method <- factor(df$Method, levels = c("Naive", "ISOMAP"))
df$Stage  <- factor(df$Stage, levels = c("Initial", "Converged"))


df %>% ggplot(aes(x=x,y=y, col=Method, linetype = Method)) + geom_path(linewidth=1) + facet_wrap(~Stage) + theme_bw() +  theme( panel.grid=element_blank()) + scale_color_manual(values=c("Naive"= "Red", "ISOMAP"="blue")) +
  theme(legend.position="bottom",
          legend.background=element_rect(fill="white",color="black"), panel.grid=element_blank(),aspect.ratio=1)
ggsave("sim_results/isomap_unwind_first_vs_converged.png", width=10, height = 5, dpi=300)

```

ISOMAP 1d 3D case


```{r}
library(plotly)

n <- 2000
t <- seq(0, 1, length.out = n)

x <- cos(2*pi*t) + 0.35*cos(6*pi*t + 0.3)
y <- sin(2*pi*t) + 0.35*sin(6*pi*t + 0.7)
z <- 0.4*sin(6*pi*t + 0.2)

fig <- plot_ly(
  x = x, y = y, z = z,
  type = "scatter3d",
  mode = "lines"
)

fig
```


# Demo Data example

## Generate data
```{r}
# Flower
dataCircle <- flower1d2D(300, 5, 0.04)$X
plot(dataCircle,asp=1)
# Half flower
dataInterval <- halfflower1d2D(300, 5, 0.04)$X
plot(dataInterval, asp=1)


```
## TDA Sim 

Homemade ggplot version


```{r}
library(TDA)




# ribbon val
runTDA_1d2D <- function(data, alpha=0.01, boot_size=30)
{
  Xlim <- c(min(data[,1]), max(data[,1]))
  Ylim <- c(min(data[,2]), max(data[,2]))
  lim <- cbind(Xlim, Ylim)
  by <- 0.01
  
  Diag <- gridDiag(data, distFct, lim = lim, by = by)
  boot <- bootstrapDiagram(X=data, FUN=distFct, lim = lim,by=by, parallel=T,alpha=alpha,B=boot_size)
  
  list(Diag=Diag, boot=boot, alpha =alpha)
}

tdaCircle <-runTDA_1d2D(dataCircle, alpha=0.01,boot_size=30)
tdaInterval <-runTDA_1d2D(dataInterval, alpha=0.01,boot_size=30)

```


# TDA Plots

```{r}

# Plot code
tda_2_plot <- function(tda_res){
  Diag <- tda_res$Diag
  boot <- tda_res$boot
  alpha <- tda_res$alpha
  
  conf_label <- paste0(round((1 - alpha) * 100), "% confidence band")
  
  pd_df <- Diag$diagram %>% unclass() %>% as.data.frame()
  
  grid_points <- seq(min(pd_df$Birth)-1, max(pd_df$Birth)+1, length.out = 100)
  
  #ribbon_df <- data.frame(
  #  x = c(grid_points - boot, rev(grid_points + boot)),
  #  y = c(grid_points + boot, rev(grid_points - boot)),
  #  Band=factor(conf_label))
  
    ribbon_df <- data.frame(
    x= c(grid_points, rev(grid_points)),
    y=c(grid_points + 2*boot,rev(grid_points)),
    Band = conf_label
  )
  
  

  
   pd_df %>%  
    ggplot() + 
    geom_point(aes(x=Birth,y=Death, color = factor(dimension), shape=factor(dimension)),size=2) +
    scale_color_manual(name="(a & d) Dimension",values=c("0"="black", "1"="red")) +
    scale_shape_manual(name="(a & d) Dimension",values = c("0" = 16, "1" = 2)) +
    geom_abline(slope=1,intercept=0)  + 
    geom_polygon(data=ribbon_df, aes(x=x,y=y, fill = Band),alpha=0.2) + 
    theme_bw() + 
    theme(panel.grid=element_blank())+ 
    scale_fill_manual(values = setNames("red", levels(ribbon_df$Band)), name = "(a & d) Band")+
     #xlim(c(0,1))+ylim(c(0,1))
      coord_cartesian(xlim = c(0, 1), ylim = c(0, 1))
}

(plotA <- tda_2_plot(tdaCircle) + ggtitle("(a)")+theme(legend.background=element_rect(fill="white",color="black")))

(plotD <- tda_2_plot(tdaInterval) + ggtitle("(d)")+theme(legend.background=element_rect(fill="white",color="black")))



```


## Show PME fits wtih different lambda values

```{r}
lambdas <- 10^c(-12, -10, -8, -6, -4, -2,0)

resA <- pmeClosed_1d_periodic(dataA, lambdas, parallel=T, optimize_lambda = F)
resB <- pmeClosed_1d_interval(dataB, lambdas, parallel=T, optimize_lambda = F)

dfA <- list_to_df(resA, "S1")
dfB <- list_to_df(resB, "interval")
  
(plotB <-df_plot_1d2D(dataA, dfA) + theme(panel.grid=element_blank()) + ggtitle("(b)")+
  geom_point(data=dfA %>% filter(lambda==1),aes(x=x,y=y, col=as.factor(lambda)),show.legend=F) + theme(aspect.ratio=1) + scale_color_discrete(name="(b & e) λ"))
  
  
(plotE <-df_plot_1d2D(dataB, dfB) + theme(aspect.ratio=1) + theme(panel.grid=element_blank()) + ggtitle("(e)")+
     theme(aspect.ratio=1)  + scale_color_discrete(name="(b & e) λ"))


```

## Functional plots

```{r}
library(scales)
res_to_alpha_df <- function(result){

  df_list <- Map(function(alpha, lam, ms) {
    
    values <- rowSums(alpha) * lam + ms
    
    data.frame(
      lambda    = lam,
      iteration = seq_len(nrow(alpha)),
      value     = values
    )
    
  },
  result$alpha_list,
  result$lambda_list,
  result$mean_sq_list)
  
  do.call(rbind, df_list)
}


alphaA <- res_to_alpha_df(resA)
alphaB <- res_to_alpha_df(resB)

(plotC <-  alphaA %>% filter(lambda==1e-8) %>% ggplot(aes(x=iteration, y = value)) + geom_point() + geom_line() + theme_bw()+ theme(panel.grid=element_blank())+ ggtitle("(c)") +
  scale_x_continuous(breaks=breaks_pretty(max(alphaA %>% filter(lambda==1e-8) %>% pull(iteration) %>% max()))) + xlab("Iteration") + ylab("Cost Functional"))

(plotF <-   alphaB %>% filter(lambda==1e-8) %>% ggplot(aes(x=iteration, y = value)) + geom_point() + geom_line() + theme_bw()+ theme(panel.grid=element_blank())+ ggtitle("(f)") +
  scale_x_continuous(breaks=breaks_pretty(max(alphaB %>% filter(lambda==1e-8) %>% pull(iteration) %>% max())))+ xlab("Iteration") + ylab("Cost Functional"))

```

## Combined results

```{r}
library(patchwork)

col1 <- plotA / plotD / guide_area()+ plot_layout(guides="collect") & theme(legend.position="bottom",
    legend.box="vertical")  
  
col2 <- (plotB+plotC) / (plotE+plotF) / guide_area() + plot_layout(guides="collect") & theme(legend.position="bottom",legend.box="vertical")

(col1 |  col2) +theme(
    plot.margin = margin(10, 10, 10, 10),   # top, right, bottom, left in pts
    legend.position = "bottom",
  )

  
```

```{r}
col1 <- plotA / plotD  + plot_layout(guides="collect") & theme(legend.position="bottom",
    legend.box="vertical")  
  
col2 <- (plotB) / (plotE)  + plot_layout(guides="collect") & theme(legend.position="bottom",legend.box="vertical")

col3 <- plotC / plotF 

plot_demo_new <- (col1 |  col2  | col3) +theme(
    plot.margin = margin(10, 10, 10, 10),   # top, right, bottom, left in pts
    legend.position = "bottom",
  )


ggsave("sim_results/demo_pic1_new.png",width = 10, height = 6, dpi = 600)
```


# Demo Monte Carlo

## Run simulation
```{r}
N <- c(100, 250, 500, 1000)



R <- 100 # 5 repetitions each
noise <- 0.04
results <- vector(mode="list", length = length(N))
names(results) <- paste0("N_",N)
lambda <- 1e-8

# Setup cluster
numCores <- parallel::detectCores() - 1
    cl <- makeCluster(numCores)
    registerDoParallel(cl)


# Begin
operator1 <- `%dopar%`
results <-operator1( foreach(i = 1:length(N), 
                              .packages = c("stats", "pracma", "foreach"),
                              .export = c("spline1d_periodic","project_grid","project_circle", "project_optimize",
                                          "spline_error","krt","krst")), {
                                                   
                                                   S <- seq(0,1,length.out=500)
                                                   # Setup Data generating curve
                                                   f_true <- flower1d2D(500,5,noise=0, t=S)$X
                                                   
                                                   df <- data.frame(t=S, x=f_true[,1],y=f_true[,2], repetition = 0)
                                                   
                                                   for(j in 1:R){
                                                     # Generate sample
                                                     data_R <- flower1d2D(N[i], 5, noise=noise)$X
                                                     spline_R <- pmeClosed_1d_periodic(data_R, lambda=lambda, 
                                                                                       parallel=F, optimize_lambda =F,
                                                                                       approx=F)
                                                     
                                                     spl <- spline_R$spline_list[[1]](S)
                                                     df <- rbind(df, data.frame(t=S, x=spl[,1], y=spl[,2],repetition=j))
                                                   }
                                                   df
                                                 }
                             
    )
    
# close cluster
 stopCluster(cl)

 
 saveRDS(results, "sim_results/monte_carlo_results.rds")
```

## Plots

Base R
```{r fig.height=4, fig.width=12}
#xlim ylim

results_N <- do.call(rbind, lapply(seq_along(results), function(i){
  df <- results[[i]]
  df$N <- N[i]
  df
})) %>% as.data.frame()

D <- max(abs(range(results_N$y)), abs(range(results_N$x)))

lims <- c(-D,D)


png("sim_results/PME_monte_carlo_star2.png", width=12, height = 4, units="in",res=600)
par(mfrow=c(1,4),oma = c(4, 0, 0, 0))
for(i in 1:length(results)){
  res <- results[[i]]
  
  true_line_data <- res %>% filter(repetition==0)
  plot(true_line_data$x, true_line_data$y, type="n", xlab="x",ylab="y",
       main=paste0("N = ", N[i]),asp=1, xlim=lims, ylim=lims)
  # add other lines
  for(j in 1:max(res$repetition)){
    line_data <- res %>% filter(repetition==j)
    lines(line_data$x, line_data$y, col=alpha("blue", alpha=0.3))
  }
  lines(true_line_data$x, true_line_data$y, type="l", col="red", lwd=2)
}
# add legend
par(xpd = NA)
mtext(side = 1, line = 1, text = "", outer = TRUE)
legend(x=-7.7, y=-2.2,
       legend = c("Sample Curves", "True Curve"),
       col = c("blue","red"),
       lty = 1,
       horiz = T,
       bty = "o")

dev.off()
```


```{r}

library(tidyverse)
library(patchwork)

plot_list <- vector(mode="list",length=length(N))
for(i in seq_along(N)){
  mydf <- results[[i]]
  # plot true
  plot_list[[i]] <-mydf %>% mutate(group = ifelse(repetition == 0, "True", "Sample")) %>% ggplot(aes(x=x,y=y, color = group)) + geom_path(linewidth=0.1, alpha=0.2) +
    scale_alpha_manual(name=NULL,values=c("True" = 1, "Sample"=0.3)) + scale_color_manual(name=NULL,values=c("True"="red", "Sample"="blue")) + theme_bw()+
    ggtitle(paste0("N = ", N[i])) +theme(panel.grid=element_blank(), 
                                         legend.background=element_rect(fill="white",color="black", linewidth=0.2),aspect.ratio = 1)

  if(c(i !=2)){
    plot_list[[i]] <- plot_list[[i]] + theme(legend.position="none") + guides(alpha="none",color="none")
  }
  
}


plot_result <- (plot_list[[1]] | plot_list[[2]] | plot_list[[3]] | plot_list[[4]]) & plot_layout(guides="collect") & theme(legend.position="bottom",
                                                                                                            plot.margin = margin(10, 10, 10, 10))


ggsave("sim_results/PME_monte_carlo_star.png", plot_result, width=10, height=6, dpi=600)
```

# Demo Lambda CV Method


# changes
1. Do not save spline lists. Just iterate.
```{r}
set.seed(1)
N <- 5000
dataCircle <- flower1d2D(N, 5, 0.04)$X

lambdas <- 10^c(seq(-12,0, length.out=100))

result_lambdas <- pmeClosed_1d_periodic(dataCircle, lambdas, optimize_lambda=T,parallel=T,lvp_grid = 500)

result_lambdas <- pmeClosed_1d_periodic_mem(dataCircle, lambdas, optimize_lambda=T,parallel=T,lvp_grid = 5*N)

rm(result_lambdas); gc()
```

Plots
```{r}
# CV vs lambda
#
plotCV<-result_lambdas$cv %>% as.data.frame() %>% ggplot(aes(x=lambda, y=cv)) + geom_point()+ scale_x_log10() + theme_bw()+
  theme(panel.grid=element_blank()) + ylab("Coefficient of Variation") + xlab("λ")+ ggtitle("(i) CV of Residuals vs λ")

plotM<-result_lambdas$vh_mean %>% as.data.frame() %>% ggplot(aes(x=lambda, y=vh_mean)) + geom_point()+ scale_x_log10() + theme_bw()+
  theme(panel.grid=element_blank()) + ylab("Mean Squared Residual") + xlab("λ")+ ggtitle("(ii) Mean of Residuals vs λ")

plotSD<- result_lambdas$vh_sd %>% as.data.frame() %>% ggplot(aes(x=lambda, y=vh_sd)) + geom_point()+ scale_x_log10() + theme_bw()+
  theme(panel.grid=element_blank())+ ylab("Standard Deviation") + xlab("λ")+ ggtitle("(iii) SD of Residuals vs λ")


S <- seq(0,1,length.out=500)
# Setup Data generating curve
f_true <- flower1d2D(500,5,noise=0, t=S)$X
df <- data.frame(t=S, x=f_true[,1],y=f_true[,2], repetition = 0)

lambda_df <- cbind(index = seq_along(lambdas), lambda=lambdas) %>% as.data.frame()
lambdasub <- lambda_df %>% filter(lambda > 10^-9 & lambda < 10^-7)

for(j in lambdasub$index){
  # Generate sample

  spl <- result_lambdas$spline_list[[j]](S)
  df <- rbind(df, data.frame(t=S, x=spl[,1], y=spl[,2],repetition=j))
}


plotFit <- df %>% mutate(group = ifelse(repetition == 0, "True", "Sample")) %>% ggplot()+  geom_point(data=data.frame(x=dataCircle[,1], y=dataCircle[,2], group="Data"),aes(x=x,y=y, col=group, alpha=group),size=0.7)+ geom_path(aes(x=x,y=y,color = group,alpha = group),linewidth=0.8) +
    scale_alpha_manual(values=c("True" = 1, "Sample"=0.3, "Data"=1)) + scale_color_manual(values=c("True"="red", "Sample"="blue","Data"="black")) + theme_bw()+
     theme(panel.grid=element_blank())+
    theme(legend.background=element_rect(fill="white",color="black")) + ggtitle("(iv) Fits in Convex Region")

```

```{r}
library(patchwork)
final_plot <-plotCV | plotM | plotSD | plotFit

ggsave("sim_results/cv_demo.png",width=24,height=6, dpi=600)
```


```{r}
set.seed(1)
N <- 1000
dataCircle <- flower1d2D(N, 5, 0.04)$X

lambdas <- 10^c(seq(-12,0, length.out=100))

result_lambdas <- pmeClosed_1d_periodic(dataCircle, lambdas, optimize_lambda=T,parallel=T, lvp_grid=5*N)

# CV vs lambda
#
```


```{r}
plotCV<-result_lambdas$cv %>% as.data.frame() %>% ggplot(aes(x=lambda, y=cv)) + geom_point()+ scale_x_log10() + theme_bw()+
  theme(panel.grid=element_blank()) + ylab("Coefficient of Variation") + xlab("λ")+ ggtitle("(i) CV of Residuals vs λ")

plotM<-result_lambdas$vh_mean %>% as.data.frame() %>% ggplot(aes(x=lambda, y=vh_mean)) + geom_point()+ scale_x_log10() + theme_bw()+
  theme(panel.grid=element_blank()) + ylab("Mean Squared Residual") + xlab("λ")+ ggtitle("(ii) Mean of Residuals vs λ")

plotSD<- result_lambdas$vh_sd %>% as.data.frame() %>% ggplot(aes(x=lambda, y=vh_sd)) + geom_point()+ scale_x_log10() + theme_bw()+
  theme(panel.grid=element_blank())+ ylab("Standard Deviation") + xlab("λ")+ ggtitle("(iii) SD of Residuals vs λ")


S <- seq(0,1,length.out=500)
# Setup Data generating curve
f_true <- flower1d2D(500,5,noise=0, t=S)$X
df <- data.frame(t=S, x=f_true[,1],y=f_true[,2], repetition = 0)

lambda_df <- cbind(index = seq_along(lambdas), lambda=lambdas) %>% as.data.frame()
lambdasub <- lambda_df %>% filter(lambda > 10^-9 & lambda < 10^-7)

for(j in lambdasub$index){
  # Generate sample

  spl <- result_lambdas$spline_list[[j]](S)
  df <- rbind(df, data.frame(t=S, x=spl[,1], y=spl[,2],repetition=j))
}


plotFit <- df %>% mutate(group = ifelse(repetition == 0, "True", "Sample")) %>% ggplot()+  geom_point(data=data.frame(x=dataCircle[,1], y=dataCircle[,2], group="Data"),aes(x=x,y=y, col=group, alpha=group),size=0.7)+ geom_path(aes(x=x,y=y,color = group,alpha = group),linewidth=0.8) +
    scale_alpha_manual(values=c("True" = 1, "Sample"=0.3, "Data"=1)) + scale_color_manual(values=c("True"="red", "Sample"="blue","Data"="black")) + theme_bw()+
     theme(panel.grid=element_blank())+
    theme(legend.background=element_rect(fill="white",color="black")) + ggtitle("(iv) Fits in Convex Region")


library(patchwork)
final_plot <-plotCV | plotM | plotSD | plotFit

ggsave("sim_results/cv_demo2.png",width=24,height=6, dpi=600)
```




